---
title: "MicroN filtering summary"
author: "Joana Séneca"
date: "4/11/2020"
output: html_document
toc: true
toc_float: true
---


## 1. Summary of this sheet
<p>This script imports, filters and exports OTU tables for the 16S and ITS1 amplicon sequencing dataset originally published by:<p>

<p>Zheng, Q. et al. Soil multifunctionality is affected by the soil environment and by microbial community composition and diversity. Soil Biol. Biochem. 136, 107521 (2019).<p>

<p>The raw fastq files are deposited on NCBI under Bioproject number PRJNA551019.<p>
<p>Shortly, both 16S and ITS datasets were:<p>
<p>1. Taxonomically filtered: OTUs not affiliated with the groups of interests were excluded;<p>
2. Abundance filtered: OTUs with an overall abundance < 0.01%  were excluded.<p>

<p>Notes: One sample (Sample_ID=SF4) had too few reads for both ITS and 16S and was probably a sequencing fail, so it was removed. This means that the Forest x Silicate treatment combination has one less replicate (n=3)<p>



## 2. Filtering ITS

### 2.1 Initial ITS phyloseq object
Taking a look at the initial phyloseq object 

```{r include=FALSE}
load("~/Desktop/MicroN/02_data/MicroN_phyloseq_objects.RData")
library(dplyr)
library(tidyverse)
library(tibble)
library(phyloseq)
```

```{r}
phy_ITS 
```

Plotting the phyloseq object colored by phylum 
```{r echo=T}
plot_bar(phy_ITS, fill='phylum_id') #there's a sample that failed, and must be removed

```

### 2.2 Removing failed sample
Removing this sample and checking the phyloseq object again:
```{r include=FALSE}
phy_ITS_pruned <- subset_samples(phy_ITS, Sequencing_ID !="SAMPLE.767.GF.II.8.60")
```
```{r}
phy_ITS_pruned
```

### 2.3 Taxonomic filtering
Check if there are unspecific taxonomic groups
```{r}
get_taxa_unique(phy_ITS_pruned, taxonomic.rank = 'phylum_id')

```

### 2.4 Abundance filtering
Since every phyla are associated with Fungi and there's nothing to exclude, we move on to the abundance filter where we'll be excluding low abundant OTUs (threshold= 0.01%)
```{r include=FALSE}
phy_ITS_f1 <- phyloseq_filter_taxa_rel_abund(phy_ITS_pruned, frac=0.0001)

```
```{r}
phy_ITS_f1
```
```{r echo=FALSE}
plot_bar(phy_ITS_f1, fill='phylum_id') 

```


### 2.5 Last checks
Let's finally check the range of reads per sample on the filtered ITS phyloseq object:
```{r}

min(sample_sums(phy_ITS_f1)) 
```
```{r}
mean(sample_sums(phy_ITS_f1))
```
```{r}
max(sample_sums(phy_ITS_f1))
```




## 3. Filtering 16S

### 3.1 Initial 16S phyloseq
Taking a look at the initial phyloseq object 

```{r include=FALSE}
load("/Desktop/MicroN/MicroN_phyloseq_objects.RData")
```

```{r}
phy_16S
```

Plotting the phyloseq object colored by phylum 
```{r echo=T}
plot_bar(phy_16S, fill='phylum_id') #there's a sample that failed and must be removed

```

### 3.2 Removing failed sample
Removing this sample and checking the phyloseq object again:
```{r include=FALSE}
phy_16S_pruned <- subset_samples(phy_16S, Sequencing_ID !="SAMPLE.325.GF.II.5.60")
```
```{r}
phy_16S_pruned
```

### 3.3 Taxonomic filtering
Check if there are unspecific taxonomic groups
```{r}
get_taxa_unique(phy_16S_pruned, taxonomic.rank = 'domain_id')

```
We will want to remove all the OTUs associated with Eukaryota and maybe also the Unclassified, but let's first take a look at how many OTUs represent these taxa, as well as their abundance across the samples

```{r echo=FALSE}
phy_16S_pruned %>% subset_taxa(domain_id %in% c("unknown", "Eukaryota")) %>% ntaxa()
```

```{r include=FALSE}
sub <- subset_taxa(phy_16S_pruned, domain_id %in% c("unknown", "Eukaryota"))
```

```{r echo=FALSE}
plot_bar(sub, fill='domain_id') # ok, they're very low abundant. Both will be removed.
```



Since they are very low abundant, we will be removing both Eukaryotes and OTUs unclassified at the domain level. Let's take a look at the new trimmed phyloseq object:

```{r include=FALSE}
phy_16S_f1 <- subset_taxa(phy_16S_pruned, !domain_id %in% c("unknown", "Eukaryota"))

```
```{r}
phy_16S_f1
```

### 3.4 Abundance filtering
Let's then apply the low abundance filter (remove OTUs with an overall abundance less than 0.01%) and take a look at the filtered object:

```{r include=FALSE}
phy_16S_f2 <- phyloseq_filter_taxa_rel_abund(phy_16S_f1, frac=0.0001)

```

```{r}
phy_16S_f2
```
```{r echo=FALSE}
plot_bar(phy_16S_f2, fill='phylum_id') 
```

### 3.5 Last checks
Let's finally check the range of reads per sample on the filtered 16S phyloseq object:
```{r}

min(sample_sums(phy_16S_f2)) 
```
```{r}
mean(sample_sums(phy_16S_f2))
```
```{r}
max(sample_sums(phy_16S_f2))
```

```{r echo=TRUE}
sessionInfo()
```

